---
title: "Exploratory Data Analysis using R"
author: "Jayendra Bhardwaj"

---



Exploratory data analysis is a process that the data analysts performs in order to discover outliers, forms of distributions and potential transformations to perform on variables in a dataset. There are many tools available, and this exercise will review some of them:

- *five number summary* 
- *histogram*
- *density plot*
- *sort plot*
- *QQ-plot*

# The 5-Number Summary

A 5-number summary is a set of 5 descriptive statistics for summarizing a numeric variable.  It consists of the variables'

* minimum
* 1st quartile
* median
* 3rd quartile
* maximum

If a distribution is skewed the mean and median will not coincide. Outliers will be detected by comparing the minimum and maximum with the quartiles.

# Getting the 5-Number Summary from R

Simple Linear Regression

```{r}
## Set the state for random number generation in R
set.seed(123)

# Pick the number of observations
n <- 100

# Pick the values for the intercept and the slope
beta_0 <- 10
beta_1 <- 2

# Assume the error has a normal distribution
# Pick the mean and standard deviation
mu <- 0
st_dev <- 0.7

# Pick the errors
err <- rnorm(n, mean = mu, sd = st_dev)

# Pick the observed inputs
x_obs <- runif(n, min = 0, max = 1)

# Generate the observed outputs
y_obs <- beta_0 + beta_1*x_obs + err

myData <- data.frame(err, x_obs, y_obs)
```

```{r}
head(myData)
```

There are two ways to generate the *5-Number Summary*. The first way is:

```{r}
fivenum(x_obs)
```

```{r}
fivenum(y_obs)
```

The second way is:

```{r}
summary(myData)
```

We see that the second method give the mean as well as the 5-Number Summery of all variables in the data frame.

We see that the the R summary function yields the 5-Number Summery, and as a bonus: the number of missing NA values, if any. And if any variable is a factor, the R summary function yields the counts of each level of the factor.

# The Histogram, Density Plot, and Sort Plot

The _histogram_ gives a view of the overall shape of the distribution of a variable and can reveal large outliers and large departures from the normal distribution. 

The _density plot_  can reveal medium outliers and clusters, while the histogram cannot because of its granularity. 

The _sort plot_ is a very good data visualization tool that reveals large and small outliers. The pattern in the gaps between adjacent values reveal clustering.

Here we overlay the curve of the normal density in the histogram plot and in the density plot; and overlay the curve of the normal quantile function in the sort plot. This allows the assessment of normality of the numeric variable.

We see that although we sample the errors from a normal distribution, the sample can have some departure from the theoretical distribution. This is expected for small sample sizes.

```{r}
attach(myData)
m <- mean(err, na.rm = TRUE)
std <- sd(err, na.rm = TRUE)
n <- 100
p <- 1:n/(n+1)
oldpar <- par(mfrow=c(1,3))
hist(
  err, 
  density=20, 
  breaks=20, 
  prob=TRUE, 
  xlab="error", 
  ylim=c(0, 0.8), 
  main="normal curve over histogram")
curve(
  dnorm(x, mean=m, sd=std),
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
plot(
  density(err),
  main = "normal curve overlay")
curve(
  dnorm(x, mean=0, sd=0.7), 
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
plot(
  p, 
  sort(err), 
  pch=".", 
  cex=2,
  main= "normal curve overlay")
curve(
  qnorm(x, mean=0, sd=0.7), 
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
detach(myData)
par(oldpar)
```

# The Normal Q-Q Plot

The Quantile-Quantile (Q-Q) Plot is used to determine how close the empirical data follow some theoretical distribution, such as the normal distribution. The plot is abbreviated as the QQ Plot. 

The QQ Plot is constructed by plotting the sorted values of a standardized variable against the associated quantiles of the standard normal distribution. It is directly related to the the sort plot and the normal quantiles. The built-in R function is `qqplot()`.

A QQ-line is drawn through the points. If the points are not close to this line, then the data may not be normal.

The QQ Plot for the `err` variable follows. We see that that the points lie very close to the diagonal line, as expected for a sample from a normal distribution.

```{r}
attach(myData)
qqnorm(
  err, 
  pch=".", 
  cex=2, 
  main="Normal Probability QQ Plot")
qqline(err)
detach(myData)
```

Sometimes it is useful to replace the sort plot with the normal QQ-plot. 

```{r}
attach(myData)
m <- mean(err, na.rm = TRUE)
std <- sd(err, na.rm = TRUE)
oldpar <- par(mfrow=c(1,3))
hist(
  err, 
  density=20, 
  breaks=20, 
  prob=TRUE, 
  xlab="error", 
  ylim=c(0, 0.8), 
  main="normal curve over histogram")
curve(
  dnorm(x, mean=m, sd=std),
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
plot(
  density(err),
  main = "normal curve overlay")
curve(
  dnorm(x, mean=0, sd=0.7), 
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
qqnorm(
  err, 
  pch=".", 
  cex=2, 
  main="Normal Probability QQ Plot")
qqline(err)
par(oldpar)
detach(myData)
```


### Exercises

Dataset ToyotaPrices

```{r, echo=FALSE}
library(readr)
setwd("/Users/jayendra/Desktop")
ToyotaPrices <- read.csv('ToyotaPrices.csv', header=T)
str(ToyotaPrices)

# Make sure to replace the path with one that works for your computer.
```


# Exercise 1. 

**Compute the *5-Number Summary* of all the vairables**

Do any of the variables exhibit some skewness? Determine if any values could be declared missing, and then convert the values to missing `NA`. Rerun the 5-Number Summary.

```{r}
summary(ToyotaPrices)

# Answer:
# Is mean and median are very unequal, it is a sign of skewness:
# Skewed: 
# Price 
# Age_08_04
# KM 
# 
# Not skewed:
# Weight 
# 
# All the rest are binary 0/1 variables, so the concenpt of skewness does not apply very well.

```

# Exercise 2. 

**Convert categorical variables to *factor***

After doing the conversions rerun the 5-Number Summary. Do any of the factor variables have "unbalanced" counts; ie, more of one kind than another? Unbalanced counts would tend to weaken the strength of a factor to predict the price of a Toyota.

```{r}
attach(ToyotaPrices)
ToyotaPrices$Automatic <- factor(Automatic)
ToyotaPrices$Mfr_Guarantee <- factor(Mfr_Guarantee)
ToyotaPrices$BOVAG_Guarantee <- factor(BOVAG_Guarantee)
ToyotaPrices$ABS <- factor(ABS)
ToyotaPrices$Airbag_1 <- factor(Airbag_1)
ToyotaPrices$Airbag_2 <- factor(Airbag_2)
ToyotaPrices$Airco <- factor(Airco)
ToyotaPrices$Automatic_airco <- factor(Automatic_airco)
ToyotaPrices$Boardcomputer <- factor(Boardcomputer)
ToyotaPrices$CD_Player <- factor(CD_Player)
ToyotaPrices$Central_Lock <- factor(Central_Lock)
ToyotaPrices$Powered_Windows <- factor(Powered_Windows)
ToyotaPrices$Power_Steering <- factor(Power_Steering)
ToyotaPrices$Radio <- factor(Radio)
ToyotaPrices$Mistlamps <- factor(Mistlamps)
ToyotaPrices$Sport_Model <- factor(Sport_Model)
ToyotaPrices$Backseat_Divider <- factor(Backseat_Divider)
ToyotaPrices$Metallic_Rim <- factor(Metallic_Rim)
ToyotaPrices$Radio_cassette <- factor(Radio_cassette)
ToyotaPrices$Tow_Bar <- factor(Tow_Bar)
detach(ToyotaPrices)
```

```{r}
summary(ToyotaPrices)

# Answer:
# 
# Unbalanced:
# BOVAG_Guarantee
# ABS
# Airbag_1
# Automatic_airco
# Boardcomputer
# CD_Player
# Power_Steering
# Radio
# Mistlamps
# Sport_Model 
# Backseat_Divider 
# Metallic_Rim
# Radio_cassette 
# Tow_Bar
```


# Exercise 3. 

**Explore the distribution of *Price***

Prepare the necessary plots, such as histogram, density plot, sort plot, QQ-plot. Is the variable normal? Is the variable skewed? Are there any clusters?

```{r}
m <- mean(ToyotaPrices$Price, na.rm = TRUE)
std <- sd(ToyotaPrices$Price, na.rm = TRUE)
n <- length(ToyotaPrices$Price)
p <- 1:n/(n+1)
oldpar <- par(mfrow=c(2,2))
hist(
  ToyotaPrices$Price, 
  density=20, 
  breaks=20,
  freq = FALSE,
  prob=TRUE, 
  xlab="ToyotaPrices$Price",
  main="normal curve over histogram")
curve(
  dnorm(x, mean=m, sd=std),
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
plot(
  density(ToyotaPrices$Price),
  main = "normal curve overlay")
curve(
  dnorm(x, mean=m, sd=std), 
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
plot(
  p, 
  sort(ToyotaPrices$Price), 
  pch=".", 
  cex=2,
  main= "sort plot w/ normal curve overlay")
curve(
  qnorm(x, mean=m, sd=std), 
  col="darkblue", 
  lwd=0.25, 
  add=TRUE)
qqnorm(
  ToyotaPrices$Price, 
  pch=".", 
  cex=2, 
  main="Normal Probability QQ Plot")
qqline(ToyotaPrices$Price)
par(oldpar)
```


# Exercise 4. 

**Scatterplot of *Price* versus *KM***

Produce the scatterplot of Price versus the number of KM (kilometers). Use both the `plot()` and the `qplot()` functions. Does the relations look like a line or a curve?

```{r}
library(ggplot2)
plot(Price ~ KM, data = ToyotaPrices)
qplot(x=KM, y=Price, data = ToyotaPrices)
# Answer: looks like a curve
```


# Exercise 5. 

**Box-Whisker plot of *Price* versus *ABS***

Produce the box-whisker plot of *Price* versus *ABS* (anti-lock breaks). Do automobiles with anti-locking breaks tend to have a higher price? Are there any outliers?

```{r}
plot(Price~ABS, data=ToyotaPrices)
qplot(x=ABS, y=Price, geom = "boxplot", data=ToyotaPrices)

# Answers: 
# Yes, prices higher with ABS
# Yes, outliers
```


# Exercise 6. 

**Correlation between *Price* and *KM***

Compute the correlation between Price and KM. Is it a strong or weak correlation? Is it positive or negative? If it is positive, what does it mean? Or- If it is negative, what does it mean?

```{r}
with(ToyotaPrices, cor(Price, KM))

# Answer:
# -0.56996
# medium negative
# As KM increase Price tends to go down, but other factors may intervene.
```









