---
title: "Hypothesis Testing using R"
author: "Jayendra Bhardwaj"
---
## Exercises

Use the uswages data in the faraway package. Make sure you identify and eliminate the missing values in the exper variable.

```{r}
# load faraway
library(faraway)

# load data
data("uswages")

# manipulating data
# we see that exper has neg. values
uswages$exper[uswages$exper <0] <-NA

# convert race, smsa, and pt to factor variables
uswages$race <- factor(uswages$race)
levels(uswages$race) <- c("White","Black")
uswages$smsa <- factor(uswages$smsa)
levels(uswages$smsa) <- c("No","Yes")
uswages$pt <- factor(uswages$pt)
levels(uswages$pt) <- c("No","Yes")

# create region, a factor variable based on the four regions ne, mw, so, we
uswages <- data.frame(uswages,
                      region =
                        1*uswages$ne +
                        2*uswages$mw +
                        3*uswages$so +
                        4*uswages$we)
uswages$region <- factor(uswages$region)
levels(uswages$region) <- c("ne","mw","so","we")

# delete the four regions ne, mw, so, we
uswages <- subset(uswages,select=-c(ne:we))

# Take care of NAs
uswages <- na.omit(uswages)

# Variable names
names(uswages)
```

### 1. Exercise

Run a model with region as predictor of wages. Show that the number of coefficients associated with region is 3.

```{r}
m.1 <- lm(wage~region, data = uswages)
coef(m.1)
```

### 2. Exercise

- Apply the `aggregate(wage ~ region, data = uswages, mean)` function in R to obtain the mean `wages` by `region`

```{r}
aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
```


```{r}
aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[2]
m.1$coef[1] + m.1$coef[2]
```

```{r}
aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[3]
m.1$coef[1] + m.1$coef[3]
```


```{r}
aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[4]
m.1$coef[1] + m.1$coef[4]
```

### 3. Exercise

- Compare the two models:
    + Model 1: wage ~ region
    + Model 2: wage ~ region + educ + exper

```{r}
m.2 <- lm(wage~region+educ+exper, data = uswages)
(tab7 <- anova(m.1, m.2))
# Answer:
# The F-Ratio 152 is big enough, so with p-value 3.03e-62 conclude that Model 2 is better, and therefore education and experience do matter.
```

- What is the conclusion - Model 1 or Model 2 is better? 

- So do education and experience matter?

### 4. Exercise

- Compare the two models:
    + Model 1: wage ~ educ + exper
    + Model 2: wage ~ region + educ + exper 

```{r}
m.1a <- lm(wage~educ+exper, data = uswages)
m.2 <- lm(wage~region+educ+exper, data = uswages)
(tab8 <- anova(m.1a, m.2))
# Answer:
# The F-Ratio 2.4 is too small, so Model 1 is better with p-value 0.066 bigger than 0.05.
# So the region of the country does not determine your wage given you already know education and experoence.
```

- The F-Ratio 2.4 is too small, so Model 1 is better with p-value 0.066 bigger than 0.05.

- Using level of significance alpha=0.05, what is the conclusion: Model 1 or Model 2 is better? 

- So does education and experience determine wage regardless of the region of the United States you live in, or does region still matter?

#5. Exercise

- Repeat exercise #4 using log(wage) for the outcome variable.

- Compare the two models:
    + Model 1: log(wage) ~ educ + exper
    + Model 2: log(wage) ~ region + educ + exper 

```{r}
m.1a <- lm(log(wage)~educ+exper, data = uswages)
m.2 <- lm(log(wage)~region+educ+exper, data = uswages)
(tab9 <- anova(m.1a, m.2))
# Answer:
# The F-Ratio 1.29 is too small, with p-value 0.28 less than 0.05.
# Conclude that region does not matter once you know education and experience.
```

- The F-Ratio 1.29 is too small, with p-value 0.28 less than 0.05.

- Using level of significance alpha=0.05, what is the conclusion: Model 1 or Model 2 is better? 

- So does education and experience determine wage regardless of the region of the United States you live in, or does `region` still matter?


```{r}
# Answers

### 1. Exercise

# Run a model with *region* as predictor of *wages*. Show that the number of coefficients associated with *region* is 3.


m.1 <- lm(wage~region, data = uswages)
coef(m.1)

### 2. Exercise

# - Apply the `aggregate(wage ~ region, data = uswages, mean)` function in R to obtain the mean `wages` by `region`
# - Show that the average wage in the *northeast* is $b_0$.

aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]

# - Show that the average wage in the *midwest* is $b_0 + b_1$ dollars.

aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[2]
m.1$coef[1] + m.1$coef[2]

# - Show that the average wage in the *south* is $b_0 + b_2$ dollars.

aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[3]
m.1$coef[1] + m.1$coef[3]

# - Show that the average wage in the *west* is $b_0 + b_3$  dollars.

aggregate(wage ~ region, data = uswages, mean)
m.1$coef[1]
m.1$coef[4]
m.1$coef[1] + m.1$coef[4]

### 3. Exercise

# - Compare the two models:
#     + Model 1: wage ~ region
#     + Model 2: wage ~ region + educ + exper

m.2 <- lm(wage~region+educ+exper, data = uswages)
(tab7 <- anova(m.1, m.2))

# Answer:
# The F-Ratio 152 is big enough, so with p-value 3.03e-62 conclude that Model 2 is better, and therefore education and experience do matter.

# - Show that the F-Ratio is 
tab7[2, 5]
# with p-value
tab7[2, 6] 

# - What is the conclusion - Model 1 or Model 2 is better? 
# - So does education and experience matter?
# - Yes they do.

### 4. Exercise

# - Compare the two models:
#     + Model 1: wage ~ educ + exper
#     + Model 2: wage ~ region + educ + exper 

m.1a <- lm(wage~educ+exper, data = uswages)
m.2 <- lm(wage~region+educ+exper, data = uswages)
(tab8 <- anova(m.1a, m.2))
# Answer:
# The F-Ratio 2.4 is too small, so Model 1 is better with p-value 0.066 bigger than 0.05.
# So the region of the country does not determine your wage given you already know education and experoence.

# - Show that the F-ratio is 
tab8[2, 5] 
# with p-value equal to 
tab8[2, 6] 
# 
# - Using level of significance $\alpha=0.05$, what is the conclusion: Model 1 or Model 2 is better? 
# 
# - So does education and experience determine wage regardless of the region of the United States you live in, or does region still matter?
# - No region does not matter with alpha=0.05. But this is a close call, since the p-value is only a little bigger than 0.05.  

### 5. Exercise

# - Repeat exercise #4 using *log(wage)* for the outcome variable.

# - Compare the two models:
#     + Model 1: log(wage) ~ educ + exper
#     + Model 2: log(wage) ~ region + educ + exper 

m.1a <- lm(log(wage)~educ+exper, data = uswages)
m.2 <- lm(log(wage)~region+educ+exper, data = uswages)
(tab9 <- anova(m.1a, m.2))
# Answer:
# The F-Ratio 1.29 is too small, with p-value 0.28 less than 0.05.
# Conclude that region does not matter once you know education and experience.

# Show that the F-ratio is 
tab9[2, 5]
# with p-value equal to 
tab9[2, 6]  

# - Using level of significance $\alpha=0.05$, what is the conclusion: Model 1 or Model 2 is better? 
# 
# - So does education and experience determine wage regardless of the region of the United States you live in, or does `region` still matter?
# - No region does not matter with alpha=0.05.
```

